Multi-stage programming with functors and monads: eliminating abstraction overhead from generic code

### 2007-07-10 SCP2008/paper/metamonads.tex
added a couple of missing refs.

In Figure~\ref{ourmonad}, the monad (yielding values of the type |v|)
is implemented as a function of two
arguments: the state (of type |s|) and the continuation. The
continuation receives the current state and the value, and
yields the answer of the type |w|.  The monad is polymorphic over the
three type parameters, which would require |monad| to be a type
constructor with three arguments. When we use this monad for code
generation, we will need yet another type variable, environment
classifier \oleg{cite Walid03?}. With type constructors taking more
and more arguments, it becomes more difficult to read and write
types -- which we will be doing extensively when writing module
signatures in Section XXX. The fact that OCaml renames all type
variables when printing out types confuses matters further. An elegant
solution to these kinds of problems has been suggested by 
Garrigue on the Caml mailing list 
(cited from
\url{http://groups.google.com/group/fa.caml/msg/e80b1245702d6b24}
no date, no exact ref). We use a single type parameter |'p| to
represent all parameters of our monad (all parameters but the type of
the monadic value |'v|). The type variable |'p| is constrained to be
the type of an object with methods (fields) |state| and |answer|. The
object may include more fields, represented by |..|. Values of that
type are not part of our computations and need not exist. We merely
use the object type, as an convenient way to specify extensible
\emph{type-level} records in OCaml.

@@

A better approach of property list has been suggested my the MLTon
team \oleg{cite MLTon}.  Encoding an object by its dual, list of
polymorphic variants.  Property list represents an extensible
object. The initial object is just the empty list. Alas, the MLton's
approach does not literally apply to us. First of all, due to
the environment classifiers, we need universes parameterized by
classifiers. Also, the property names are generated automatically
in MLton approach, based on generativity of exceptions or reference
cells. That means if some particular aspect happens to be included
twice, the state will be incompatible. We will have to worry about
sharing or the lack of sharing of our aspects. We would rather
prefer our aspects to be stateless and work the same way whether two
instances of the aspect are shared or not. Fortunately, OCaml offers a
way to build open unions using polymorphic variants. Each variant is
identified by a manifest tag, e.g., `Tdet, and may include a
value. The tags are not generative and provide the form of manifest
naming (somewhat similar to symbols of Lisp and Scheme).

@@

\section{Runge-Kutta solvers}
\label{s:ode}
\oleg{Drop this? no space}
\jacques{We should drop this as a whole section, but I still think we 
should mention it as something that we have done.  Mostly to say that
things ``extend'' to other families too, and that no new difficulties 
were encountered.}

>>>>

In Figure~\ref{ourmonad}, the monad (yielding values of the type |v|)
is implemented as a function of two
arguments: the state (of type |s|) and the continuation. The
continuation receives the current state and the value, and
yields the answer of the type |w|.  The monad is polymorphic over the
three type parameters, which would require |monad| to be a type
constructor with three arguments. When we use this monad for code
generation, we will need yet another type variable, environment
classifier \cite{taha-environment}. With type constructors taking more
and more arguments, it becomes more difficult to read and write
types -- which we will be doing extensively when writing module
signatures in Section XXX. The fact that OCaml renames all type
variables when printing out types confuses matters further. An elegant
solution to these kinds of problems has been suggested by 
Garrigue on the Caml mailing list 
(cited from
\url{http://groups.google.com/group/fa.caml/msg/e80b1245702d6b24}
no date, no exact ref). We use a single type parameter |'p| to
represent all parameters of our monad (all parameters but the type of
the monadic value |'v|). The type variable |'p| is constrained to be
the type of an object with methods (fields) |state| and |answer|. The
object may include more fields, represented by |..|. Values of that
type are not part of our computations and need not exist. We merely
use the object type, as an convenient way to specify extensible
\emph{type-level} records in OCaml.

@@

A better approach of property list has been suggested my the MLton
team \cite{mlton-proplist}.  Encoding an object by its dual, list of
polymorphic variants.  Property list represents an extensible
object. The initial object is just the empty list. Alas, the MLton's
approach does not literally apply to us. First of all, due to
the environment classifiers, we need universes parameterized by
classifiers. Also, the property names are generated automatically
in MLton approach, based on generativity of exceptions or reference
cells. That means if some particular aspect happens to be included
twice, the state will be incompatible. We will have to worry about
sharing or the lack of sharing of our aspects. We would rather
prefer our aspects to be stateless and work the same way whether two
instances of the aspect are shared or not. Fortunately, OCaml offers a
way to build open unions using polymorphic variants. Each variant is
identified by a manifest tag, e.g., `Tdet, and may include a
value. The tags are not generative and provide the form of manifest
naming (somewhat similar to symbols of Lisp and Scheme).

@@

\section{Runge-Kutta solvers}
\label{s:ode}
\oleg{Drop this? no space}
\jacques{We should drop this as a whole section, but I still think we 
should mention it as something that we have done.  Mostly to say that
things ``extend'' to other families too, and that no new difficulties 
were encountered.}
\oleg{Ok, so we mention Runge-Kutta only in the introduction, and drop
  this section then, right? Do you want to do that?}
